---
title: Many models, same data
author: Lydia English and Gina Nichols
date: '2020-04-22'
slug: many-models
categories: 
  - Weeds
  - Model selection
tags: 
  - r
  - Grad School
subtitle: ''
summary: ''
authors: []
lastmod: '2020-04-23T14:27:24-05:00'
draft: false
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

### Introduction

Let's talk about model selection! As a non-statistician, I find the process of choosing an appropriate statistical model to be slightly intimidating, even agonizing at times. In the hopes of demystifying this process for other non-statisticians, this post attempts to walk you through how [Gina Nichols](https://vanichols.netlify.app/) and I decided on the appropriate models and stats for an upcoming manuscript. Constructive feedback is always welcome.

Gina and I are using data from [this project](/project/pfiweeds/). In very brief terms we're interested in the effect of cover crops on the weed seed bank of corn/soybean cropping systems. While cover crops are oft cited as a practice than can help decrease erosion or nutrient export from agricultural fields, we are curious if using cover crops can help decrease the weed seed bank. This is especially important because of the increasing prevalence of herbicide resistant weeds across the Midwest (and beyond). 

### Data

The data can be accessed from Gina Nichols' Github repository like so:

`remotes::install_github("vanichols/PFIweeds2020")`

The raw data are in the data-frame `pfi_ghobsraw`. These are observations from soil collected at each site which Gina then incubated in a greenhouse to grow weeds. We don't need to clean the data very much, but we will do a couple steps to consolidate observations. 

A couple other notes on the experimental layout:

* There are four `site_sys`, which encompass three different farms and two different cropping systems
  + Boyd Grain
  + Boyd Silage
  + Funcke Grain
  + Stout Grain

<br>
  _FYI: Grain systems employ a two year rotation of corn and soybeans and sell both crops off as feed grain, ethanol grain, etc. The silage system also has a 2 year rotation of corn and soybeans, but in the corn year the crop is harvested earlier (while it's still green) and sold as silage (fermented cattle feed)_
  
* There are two cover crop treatments `cc_trt` = `none` and `cc_rye`= `rye`
* There are 4-5 blocks per treatment, which will be our random effect. 

### Library loading & initial data visualization

```{r setup, include = TRUE, warning = FALSE, message = FALSE}
library(dplyr)
library(lme4)
library(lmerTest)
library(ggResidpanel)
library(performance)
library(ggplot2)
library(patchwork)
library(PFIweeds2020)
theme_set(theme_bw())

```

A couple notes about the libraries we used:

* We are fitting models using [lme4](https://arxiv.org/abs/1406.5823) and [lmerTest](https://backend.orbit.dtu.dk/ws/portalfiles/portal/140635100/lmerTestJStatSoft2017.pdf) packages.
* We are using two different packages to examine model fit/diagnostics.
    + [ggResidpanel](https://CRAN.R-project.org/package=ggResidpanel) has much improved versions of model diagnostic plots. Would highly recommend!
    + [performance](https://CRAN.R-project.org/package=performance) also has improved model diagnostic plots as well as simple ways of calculating R<sup>2</sup> for fixed & random effects and checking for over-dispersion.


```{r}
data("pfi_ghobsraw")

# Quick data cleaning
dat <- 
  pfi_ghobsraw %>% 
  dplyr::mutate_if(is.numeric, tidyr::replace_na, 0) %>%
  dplyr::group_by(site_name, field, sys_trt, cc_trt, rep, blockID) %>%
  dplyr::summarise_if(is.numeric, sum) %>%
  dplyr::ungroup() %>%
# Sum all seeds by species  
  dplyr::mutate(totseeds = rowSums(.[, 7:24])) %>%
  tidyr::unite(site_name, sys_trt, col = "site_sys", remove = T) %>% 
  select(-field, -rep, -c(AMATU:UB)) %>%
  mutate(cc_trt = recode(cc_trt,         
                          no = "none",
                          rye = "cc_rye"))
head(dat)
```

We are interested in `totseeds` (total seeds) as a response variable, which is an integer value. In all of our mixed models our fixed effects are `site_sys` and `cc_trt` and our random effect is `blockID`.

But before we get to modeling, let's just quickly look at the data and see what we are working with. 

```{r, echo = FALSE}
dat %>%
  ggplot(aes(site_sys, totseeds))+
  geom_boxplot(aes(color = cc_trt))+
  geom_point(aes(color = cc_trt), position = position_dodge(width = 0.75))+
  labs(y = NULL)

```
There is one point that is much higher than the rest, in the `Funcke_grain` cover crop treatment, which may be a little cause for concern...we're going to make another data-set without it. 

```{r}
dat2 <- dat %>%
  filter(totseeds < 700)

```

Let's also plot the distribution of `totseeds`. It is pretty skewed. If we log-transform the raw data we get a much nicer distribution and our large values are much less far apart than our small values. 

```{r, echo = FALSE}
p1 <- 
  dat2 %>%
  ggplot(aes(totseeds)) +
  geom_histogram(binwidth = 40, fill = "white", color = "black")+
  ggtitle("Raw Data")+
   theme(plot.title.position = "plot")

p2 <- 
  dat2 %>%
  ggplot(aes(log(totseeds)))+
  geom_histogram(binwidth = 0.5, fill = "white", color = "black")+
  ggtitle("Log-transformed Data")+
  theme(plot.title.position = "plot")

p1 + p2

```

### Model fitting

OK now it's time to start fitting out models. Given the graphs above, our first thought was to run a `lmer` with a log-transformed response.

(Here we are also examining whether there is any interaction between `site_sys` and `cc_trt`, and looks like there is. We are therefore going to keep the interaction in all the models going forward) 

```{r, message = FALSE}

m1 <- lmer(log(totseeds) ~ site_sys * cc_trt + (1|blockID), data = dat2)
m2 <- lmer(log(totseeds) ~ site_sys + cc_trt + (1|blockID), data = dat2)

anova(m1, m2) # keep interaction
```

Before we dive into the summary values and statistical reporting lets check out our diagnostic plots:

```{r}
ggResidpanel::resid_panel(m1)
```

This doesn't look bad, but perhaps it could look better...

To stick within the normal-distribution framework we could also try another transformation. The square-root transformation can also work for count data. 

```{r}
dat %>%
  ggplot(aes(sqrt(totseeds)))+
  geom_histogram(binwidth = 1, fill = "white", color = "black")+
  ggtitle("Sqrt-transformed Data")+
  theme(plot.title.position = "plot")
```

The data still look a little skewed, but that's OK...

```{r}
m1b <- lmer(sqrt(totseeds) ~ site_sys * cc_trt + (1|blockID), dat2)
resid_panel(m1b)
```

These diagnostic plots look pretty similar, although the square root transformation might be a little better. The problem with the square root transformation is that it renders the data pretty uninterpretable and we'd like to be able to interpret our model estimates. 

Since we are working with count data, however, we have other options in our stats toolbox - generalized linear models. We don't even need to load another package to do this, we just need to call `glmer` from the {lme4} package and specify which distribution we're fitting.

Our first attempt with `glmer` will be a Poisson distribution, which is meant for count data and often (but not always) uses a log link function. 

```{r}
p1a <- glmer(totseeds ~ site_sys*cc_trt + (1|blockID), data = dat2, 
            family = poisson(link = "log"))
```

It looks like our model successfully fit the data. But we should always check for over-dispersion in our Poisson models, which is when the variance >> mean. 

```{r}
performance::check_overdispersion(p1a) 
```


Just as an aside, you can also calculate dispersion by hand using one of these two methods:
```{r}
# option 1
1 - pchisq(deviance(p1a), df.residual(p1a)) 
# option 2
r <- resid(p1a,type="pearson")
1-pchisq(sum(r^2),df.residual(p1a))

```
Either way our p-value is very close to 0, so it looks like our data are over-dispersed.

Now we have two options if we want to continue down the `glmer` path. We could:

1. Add an observation-level random effect to our Poisson model.

2. Switch to a different distribution, like the negative binomial.

Let's start with option 1 and add an observation-level random effect. To learn more about why this helps, see [this article](https://peerj.com/articles/616/).

```{r}
dat2 <- 
  dat2 %>%
  mutate(obs_id = paste("obs", 1:n(), sep = "_"))

p1b <- glmer(totseeds ~ site_sys*cc_trt + (1|blockID) + (1|obs_id), data = dat2, 
            family = poisson(link = "log"))
```

```{r}
performance::check_overdispersion(p1b)

```

Adding our observation-level random effect fixed out over-dispersion problem. Great! 

But let's also try option 2 and fit the data to a negative-binomial model.

```{r}
# option 2, negative binomial....
g1a <- glmer.nb(totseeds ~ site_sys*cc_trt + (1|blockID), data = dat2)

```

This model also converges, which is great! In my limited experience using negative binomial models this isn't always true so this is a small victory.

Let's now compare the summary values of the two models:

```{r}
summary(p1b)
```

```{r}
summary(g1a)
```

```{r}
performance::compare_performance(p1b, g1a)
```

Both these models have very similar AICs, and they also have very similar estimates and p-values which is reassuring, since our interpretation of the data won't change depending on the model we choose. 

In the end we decided to go with the Poisson-observation-level-random-effect model (named here `p1b`). Poisson models are slightly more simple than negative binomial and built for count data. 


:::summary
 Recap of our models: 

i. Log-transformation: decent fit but slight loss in interpretability
ii. Square-root transformation: slightly better fit (at least visually) but complete loss of interpretability
iii. Poisson generalized linear model: works but over-dispersed
iv. Poisson generalized linear model with observation-level random effect: fixes overdispersion and seems to fit well! 
v. Negative binomial generalized linear model: fits very similarly to Poisson, but slightly more complicated model. 
:::